The CMS Data Aggregation System

### 2010-05-04 doc/paper.tex
Added Simon's changes for sections 4.3-4.6
git-svn-id: svn+ssh://svn.cern.ch/reps/CMSDMWM/DAS/trunk@1845 4525493e-7705-40b1-a816-d608a930855b

\subsection{DAS Mapping DB\label{MappingDB}}
The Mapping DB represents DAS {\it registration} service. It collects
information about data-service APIs, translates API input/output
parameters into DAS keys and contains mapping between DAS records
and their UI representation.

%The Mapping DB hold the mapping between data-service APIs and DAS keys exposed to
%the users. It created a necessary glue between data-services and identifies
%how their data can be related to each other. Basically Mapping DB was positioned 
%as a {\it registration} service with ability to translate notations among data-services.
Upon identification of APIs which participate in DAS 
we stored their names, input parameters and associated DAS keys
into the Mapping DB. For example
\begin{verbatim}
{"api":{"params":{"name":""}, "name":"CMSNametoSE"}, "system":"sitedb",
 "api2das": [{"pattern": "re.compile('^T[0-3]_')", 
              "api_param": "name", "das_key": "site"}], 
 "daskeys": [{"pattern": "re.compile('^T[0-3]_')", 
              "key": "site", "map": "site.name"}]}
\end{verbatim}
represents DAS mapping record.\footnote{
The data-service URLs were kept in the configuration file.
}
This record can be read as following:
the \verb+CMSNametoSE+ API provided by 
\verb+sitedb+ data-service accepts \verb+name+ as input parameter. It
confirms regular expression pattern, \verb+re.compile('^T[0-3]_')+, and
corresponds to \verb+site+ DAS key. The \verb+site+ key
maps into \verb+site.name+ within DAS record.

Therefore the Mapping DB identifies a set of pre-defined DAS keys, 
e.g. \verb+site, file, run+, etc., to be used by end-users in their queries.
Each key was mapped into set of APIs whose output was aggregated afterwards.

\subsection{DAS Analytics DB}
The DAS Analytics DB collected information
about user requests placed to the system. Each request was recorded. Upon its
decomposition into set of selection keys and conditions we also recorded
API name and input parameters into the Analytics DB and 
kept updating their counter information for repeated calls. 
This allowed to keep track of frequency of API calls and establish 
pre-fetch strategies for most common queries.

\subsection{DAS caching system}
The DAS caching system was used to dynamically fetch
and aggregate data upon user requests. We used the following 
algorithm (code snippets are written in Python \cite{Python}):
\begin{enumerate}[1.]
\item Decompose query into set of conditions (\verb+conds+) 
and selection keys (\verb+skeys+).
\item Check cache for query hash. If found, look-up results from the cache,
otherwise proceed to the next step.
\item Identify list of APIs which can serve requested data
\begin{verbatim}
apilist = [r for r in mapping_db(skeys, conds)]
\end{verbatim}
Here \verb+r+ represents a triplet \verb+url, api, args+ and 
\verb+mappind_db+ defines translation between input query and data-service
APIs which can delivery the data.
\item Iterate over APIs
\begin{verbatim}
for url, api, args in apilist:
    if  superset_in_analytics(api, args):
        continue
    for row in parser(get_data(url, api, args)):
        yield row
\end{verbatim}
Here \verb+superset_in_analytics+ checks if provided \verb+api, args+
are subset of already registered in Analytics DB api call(s). The 
\verb+get_data+ and \verb+parser+ are define methods to fetch and parse
data, respectively.
\item Perform aggregation by iterating over records from the previous step
\begin{verbatim}
if cache_has(daskey):
   existing_record = get_from_cache(daskey)
   merge(row, existing_record)
   insert_into_cache(row)
   delete_from_cache(existing_record)
else:
   insert_into_cache(row)
\end{verbatim}
The \verb+daskey+
is a key which identifies the record defined in Mapping DB, e.g. \verb+file.name+
for \verb+file+ record.
We did use a bulk insert operations at this step to increase throughput. 
\item Get results from the cache
\end{enumerate}
We also had an option to run independent set of daemons, DAS robots, who
monitor Analytics DB for most popular requests and pre-fetch the most
common data into the cache.

\subsection{DAS queries}
All DAS queries were expressed in a free text-based form, either as a 
set of keywords or key-value pairs, where a pair can represent
condition, e.g. \verb+key > value+. We used a pre-defined set of keys registered 
in DAS to describe common entities of our data. For example, \verb+dataset+, 
\verb+file+, \verb+block+, \verb+run+, \verb+event+, etc. Their choice
was based on the analysis of daily work operations of our users, see \cite{DBS07}.
Each key represents a DAS record. Every record was stored into the cache
in JSON format. Due to the ``schemaless'' nature of underlying back-end (MongoDB)
we were able to store DAS records of arbitrary structure, e.g.
dictionary, lists, key-value pairs, etc. Therefore every DAS key had a 
set of attributes representing JSON structure. For instance a document
\begin{verbatim}
{"block": {"name": "abc", "size": 1, "replica": [...]}}
\end{verbatim}
associated with a \verb+block+ key who had the following attributes
\verb+name+, \verb+size+, \verb+replica+. The set of attributes is based 
on a set of meta-data information stored in a record and therefore
a nested structures were allowed, e.g. \verb+block.replica.size+. Our users 
where able to specify a context of their queries by using appropriate 
attribute(s). For example, query \verb+block+ represents search for all
block records rather then ``block'' string, while query
\verb+block.name=abc*+ instructs the system to look-up appropriate blocks
with matched names. The numerical values, conditions, groupping were
easy to express too, e.g. \verb+run > 10 and run < 100+ query 
was used to get all run records for provided range of run numbers. 
To complement the strength of the queries, we do plan to index 
dynamically DAS records and add a keyword-based 
search within record content later.

All input queries were transformed into MongoDB query syntax who is itself
a JSON record. It contains a set of selection 
keys (\verb+fields+) and conditions (\verb+spec+). For example, a query
\verb+dataset, file=abc, run>10+
was represented as
\begin{verbatim}
{"fields": ["dataset"], "spec": {"file": "abc", "run": {"$gt": 10}}}
\end{verbatim}
Therefore users queries were stored into Analytics DB similar to other DAS 
records and used for further analysis.
%Such syntax was perfect fit for us, since it allows to store all queries
%as JSON documents into Analytics DB for further analysis. It is worthwhile to note that
%MongoDB query syntax is very reach. It supports advanced queries with 
%boolean expressions, conditional operators, regular expressions, etc. providing
%a real power for our end-users.

\subsection{Data-services and aggregation}
As was shown in section \ref{MappingDB} the DAS Mapping DB became an authorative
source of information about all data-services participating in DAS.
Based on the information stored in Mapping DB we were able to 
retrieve appropriate data from different data-services, re-map them into 
DAS notations and aggregate them on demand. For instance, the Data 
Bookkeeping System (DBS) \cite{DBS} and data location system (PhEDEx) \cite{PhEDEx}
were provided information about CMS blocks. In former case, DBS 
stored information about blocks and their relationships to other 
physics objects, such as dataset and files, while PhEDEx kept
information about its location and replicas. Upon a user request to 
find a block meta-data, we were able to identify appropriate set of 
DBS and PhEDEx APIs, retrieve data, re-format their output 
into DAS notations and store results into DAS cache. The records
were merged (aggregated) based on identical key-value pairs, e.g.
based on \verb+block.name+ key. Here is an example of such aggregated record:
\begin{verbatim}
{"das_id": ["4b195404e2194e21c4000030", "4b1953fee2194e21c4000002"], 
 "block": [{"name": "/a/b/c#123", "replica": [{"complete": "y", 
            "subscribed": "n", "custodial": "n", "site": "T1_US_Buffer", 
            "nfiles": "18", "se": "srm.site.com", 
            "size": "112694171838"}, {....}]},
           {"name": "a/b/c#123", "nevents": "570200", 
            "created_by": "USER_DN", "path": "/a/b/c", ...}],
 "das": [{"expire": 1259952908}, {"expire": 1259954702}]}
\end{verbatim}
The \verb+das_id+ part refers to api call records stored in Analytics DB,
the \verb+block+ part represents an aggregation block meta-data information from DBS
and PhEDEx systems and the \verb+das+ part contains their expiration time stamps.
Usage of this information was valuable contribution in debugging process of
data-services themselves, e.g. identification of data-service inconsistencies, 
latency studies, etc.

>>>>

\subsection{DAS Mapping DB\label{MappingDB}}
The Mapping DB is used to track services known to DAS and how to interact with them.
%The Mapping DB represents DAS {\it registration} service. 
It collects
information about data-service APIs, translates API input/output
parameters into DAS keys and contains mapping between DAS records
and their UI representation.

On identifying API's that are accessible to DAS we store
their names, input parameters and associated DAS keys
into the Mapping DB. For example
\begin{verbatim}
{"api":{"params":{"name":""}, "name":"CMSNametoSE"}, "system":"sitedb",
 "api2das": [{"pattern": "re.compile('^T[0-3]_')", 
              "api_param": "name", "das_key": "site"}], 
 "daskeys": [{"pattern": "re.compile('^T[0-3]_')", 
              "key": "site", "map": "site.name"}]}
\end{verbatim}
represents DAS mapping record.\footnote{
The data-service URL's are kept in the configuration file.
}
This record can be read as following:
the \verb+CMSNametoSE+ API provided by 
\verb+sitedb+ data-service accepts \verb+name+ as input parameter. It
confirms regular expression pattern, \verb+re.compile('^T[0-3]_')+, and
corresponds to \verb+site+ DAS key. The \verb+site+ key
maps into \verb+site.name+ within DAS record.

Therefore the Mapping DB identifies a set of pre-defined DAS keys, 
e.g. \verb+site, file, run+, etc., to be used by end-users in their queries.
Each key is mapped into set of API calls, whose output is retrieved,
cached and aggregated as required.

\subsection{DAS Analytics DB}
The DAS Analytics DB collects information on user requests against the system.
Each request that DAS handles is recorded.
The Analytics DB tracks:
\begin{itemize}
\item The users request
\item How DAS has mapped the request onto data service API's
\item The time taken by the remote data services to process the API calls
\end{itemize}
By recording this information we can plan pre-fetch strategies for common queries,
identify issues in remote data services and cross check that DAS resolves requests
in a deterministic manner.

% Upon its
%decomposition into set of selection keys and conditions we also recorded
%API name and input parameters into the Analytics DB and 
%kept updating their counter information for repeated calls. 
%This allowed to keep track of frequency of API calls and establish 
%pre-fetch strategies for most common queries.

\subsection{DAS caching system}
The DAS caching system is used to dynamically fetch
and aggregate data upon user requests. We use the following 
algorithm (code snippets are written in Python \cite{Python}):
\begin{enumerate}[1.]
\item Decompose query into set of conditions (\verb+conds+) 
and selection keys (\verb+skeys+).
\item Check cache for query hash. If found, look-up results from the cache,
otherwise proceed to the next step.
\item Identify list of APIs which can serve requested data
\begin{verbatim}
apilist = [r for r in mapping_db(skeys, conds)]
\end{verbatim}
Here \verb+r+ represents a triplet \verb+url, api, args+ and 
\verb+mappind_db+ defines translation between input query and data-service
APIs which can delivery the data.
\item Iterate over APIs
\begin{verbatim}
for url, api, args in apilist:
    if  superset_in_analytics(api, args):
        continue
    for row in parser(get_data(url, api, args)):
        yield row
\end{verbatim}
Here \verb+superset_in_analytics+ checks if provided \verb+api, args+
are subset of already registered in Analytics DB api call(s). The 
\verb+get_data+ and \verb+parser+ are define methods to fetch and parse
data, respectively.
\item Perform aggregation by iterating over records from the previous step
\begin{verbatim}
if cache_has(daskey):
   existing_record = get_from_cache(daskey)
   merge(row, existing_record)
   insert_into_cache(row)
   delete_from_cache(existing_record)
else:
   insert_into_cache(row)
\end{verbatim}
The \verb+daskey+
is a key which identifies the record defined in Mapping DB, e.g. \verb+file.name+
for \verb+file+ record.
We did use a bulk insert operations at this step to increase throughput. 
\item Get results from the cache
\end{enumerate}
We also have an option to run independent set of daemons, DAS robots, who
monitor Analytics DB for most popular requests and pre-fetch the most
common data into the cache.

\subsection{DAS queries}
All DAS queries are expressed in a free text-based form, either as a 
set of keywords or key-value pairs, where a pair can represent
condition, e.g. \verb+key > value+. We use a pre-defined set of keys registered 
in DAS to describe common entities of our data. For example, \verb+dataset+, 
\verb+file+, \verb+block+, \verb+run+, \verb+event+, etc. Their choice
is based on the analysis of daily work operations of our users, see \cite{DBS07}.
Each key represents a DAS record. Every record is stored into the cache
in JSON format. Due to the ``schemaless'' nature of underlying back-end (MongoDB)
we are able to store DAS records of arbitrary structure, e.g.
dictionary, lists, key-value pairs, etc. Therefore every DAS key has a 
set of attributes representing JSON structure. For instance a document
\begin{verbatim}
{"block": {"name": "abc", "size": 1, "replica": [...]}}
\end{verbatim}
is associated with a \verb+block+ key who has the following attributes
\verb+name+, \verb+size+, \verb+replica+. The set of attributes is based 
on a set of meta-data information stored in a record and therefore
a nested structures are allowed, e.g. \verb+block.replica.size+. Our users 
are able to specify a context of their queries by using appropriate 
attribute(s). For example, query \verb+block+ represents search for all
block records rather then ``block'' string, while query
\verb+block.name=abc*+ instructs the system to look-up appropriate blocks
with matched names. The numerical values, conditions, grouping are
easy to express too, e.g. \verb+run > 10 and run < 100+ query 
is used to get all run records for provided range of run numbers. 
To complement the strength of the queries, we do plan to index 
dynamically DAS records and add a keyword-based 
search within record content later.

All input queries are transformed into MongoDB query syntax who is itself
a JSON record. It contains a set of selection 
keys (\verb+fields+) and conditions (\verb+spec+). For example, a query
\verb+dataset, file=abc, run>10+
is represented as
\begin{verbatim}
{"fields": ["dataset"], "spec": {"file": "abc", "run": {"$gt": 10}}}
\end{verbatim}
Therefore users queries are stored into Analytics DB similar to other DAS 
records and used for further analysis.

\subsection{Data-services and aggregation}
As was shown in section \ref{MappingDB} the DAS Mapping DB became an authorative
source of information about all data-services participating in DAS.
Based on the information stored in Mapping DB we were able to 
retrieve appropriate data from different data-services, re-map them into 
DAS notations and aggregate them on demand. For instance, the Data 
Bookkeeping System (DBS) \cite{DBS} and data location system (PhEDEx) \cite{PhEDEx}
provide information about CMS blocks. In former case, DBS 
stored information about blocks and their relationships to other 
physics objects, such as dataset and files, while PhEDEx kept
information about its location and replicas.

Upon a user request to 
find a block meta-data, we were able to identify appropriate set of 
DBS and PhEDEx APIs, retrieve data, re-format their output 
into DAS notations and store results into DAS cache. The records
were merged (aggregated) based on identical key-value pairs, e.g.
based on \verb+block.name+ key. Here is an example of such an aggregated record:
\begin{verbatim}
{"das_id": ["4b195404e2194e21c4000030", "4b1953fee2194e21c4000002"], 
 "block": [{"name": "/a/b/c#123", "replica": [{"complete": "y", 
            "subscribed": "n", "custodial": "n", "site": "T1_US_Buffer", 
            "nfiles": "18", "se": "srm.site.com", 
            "size": "112694171838"}, {....}]},
           {"name": "a/b/c#123", "nevents": "570200", 
            "created_by": "USER_DN", "path": "/a/b/c", ...}],
 "das": [{"expire": 1259952908}, {"expire": 1259954702}]}
\end{verbatim}
The \verb+das_id+ refers to api call records stored in Analytics DB,
the \verb+block+ part represents an aggregation block meta-data information from DBS
and PhEDEx systems and the \verb+das+ part contains their expiration time stamps.
Usage of this information was valuable contribution in debugging process of
data-services themselves, e.g. identification of data-service inconsistencies, 
latency studies, etc.
